Data migration under a Changing Schema

### 2023-03-19 2022 Migration/articleMigration.tex

   Relations serve to organize and store data, to allow a developer to represent facts.
   In this paper, variables $r$, $s$, and $d$ represent relations.
   All relations are taken from an infinite set $\Rels$.
   $\Rels$ is disjoint from $\Concepts$ and $\Atoms$.
   Every relation $r$ has a name, a source concept, and a target concept.
   The notation $r=\declare{n}{A}{B}$ denotes that relation $r$ has name $n$, source concept $A$, and target concept $B$.

@@

   A relation $r$ can serve as a container of pairs,
   as defined by the function $\id{pop}_r:\Dataset\rightarrow\powerset{\Pair{\Atoms}{\Atoms}}$.
   It defines a set of pairs, also known as the population of $r$:
\begin{equation}
   \pop{r}{\dataset}\ =\ \{ \pair{a}{b}\mid\ \triple{a}{r}{b}\in\triples_{\dataset}\}
\label{eqn:pop}
\end{equation}
   Equation~\ref{eqn:wellTypedEdge} implies that for every dataset $\dataset$:
\[\pair{a}{b}\in\pop{\declare{n}{A}{B}}{\dataset}\ \Rightarrow\ a\inst A\ \wedge\ b\inst B\]
   For a developer, this means that the type of an atom depends only on the relation in which it resides; not on the actual population of the database.
   This allows for static typing, which has well established advantages for the software engineering process~\cite{HanenbergKRTS14,Petersen2014}.

@@

   A schema must ensure that triples ``make sense'' in the semantics defined by the schema.
   This means that it must satisfy the following requirements.
\begin{eqnarray}
   \declare{n}{A}{B}\in\rels&\Rightarrow&A\in\concepts\ \wedge\ B\in\concepts
   \label{eqn:relationsIntroduceConcepts}\\
   A\ \isa\ B&\Rightarrow&A\in\concepts\ \wedge\ B\in\concepts
   \label{eqn:isasIntroduceConcepts}\\
   \isa&&\text{is a partial order on $\concepts$}
   \label{eqn:isasPartialOrder}\\
   \forall u\in\rules,\exists\dataset\in\Dataset-\emptyset&:&\sat{u}{\dataset}
   \label{eqn:trueifyableRules}
\end{eqnarray}
   Note that these requirements do not depend on any particular dataset,
   so they can be checked statically.
   The type system of Ampersand~\cite{vdWoude2011} ensures that every information system it generates satisfies these requirements.

@@

   The information system enforces the semantics by ensuring that the following requirements remain true at all times:
\begin{eqnarray}
   \triple{a}{\declare{n}{A}{B}}{b}\in\triples&\Rightarrow&\declare{n}{A}{B}\in\rels
   \label{eqn:define R}\\
   \forall u\in\rules&:&\sat{u}{\dataset}
   \label{eqn:satisfaction}
\end{eqnarray}
   Requirement~\ref{eqn:define R} ensures that the information system only works with triples whose relation is defined in the schema.
   Requirement~\ref{eqn:satisfaction} ensures that the dataset satisfies all rules in the schema.
   The rules in the schema serve as invariants,
   i.e.\ constraints that the data in the database must satisfy at all times.
   At times in which the dataset does not satisfy a rule, that rule is said to be {\em broken}.
   In those cases, something or someone must take action to fix the situation and restore invariance of that rule.
   The system is {\em at rest} when every rule in the schema is satisfied.

@@

\subsection{Changing data}
\label{sct:Changing data}
   Events can cause a dataset to change.
   So, we model events as functions $\Dataset\rightarrow\Dataset$.
   We will use variable $e$, $e'$, $e''$, and so on to represent events.
   We distinguish just two types of event: insert and delete events.
   An event is a label (\id{ins} or \id{del}) with a dataset that is to be inserted or deleted.
   If $\triples$ is the set of triples inserted or deleted by an event $e$,
   we denote that event either by $\ins{\triples}$ or $\del{\triples}$.
   They are defined by:
\begin{eqnarray}
   e(\dataset)&=&\id{case}\ e\ \id{of}\\
   &&\begin{array}[t]{@{}rcl}
      \ins{\triples}&=&\pair{\triples_{\dataset}\cup\triples}{\id{insInst}_{\triples}(\inst_{\dataset})}\\
      \del{\triples}&=&\pair{\triples_{\dataset}-\triples}{\id{delInst}_{\triples}(\inst_{\dataset})}
   \end{array}
\end{eqnarray}
   in which $\id{insInst}_{\triples}$ transforms $\inst_{\dataset}$ such that $\ins{\triples}(\dataset)$ complies with
   equations~\ref{eqn:specialization} and~\ref{eqn:wellTypedEdge},
   and likewise for $\id{delInst}_{\triples}$.

   As the system registers events, it inserts or deletes triples in its dataset.
   This may violate requirement~\ref{eqn:satisfaction} when one or more rules are broken,
   i.e.\ if there is a rule $u$ for which $\sat{u}{e(\dataset)}$ does not hold.
   In that case, requirement~\ref{eqn:satisfaction} prohibits the system to be in state $e(\dataset)$,
   so either the system stays in the state $\dataset$ or there is a reaction $e'$ that restores invariance of the broken rules:
\begin{equation}
   \sat{u}{\dataset}\Rightarrow\sat{u}{e'(e(\dataset))}
   \label{eqn:restoring}
\end{equation}
   The combination of events $e$ and $e'$ is called a {\em transaction}.

   In practice, there are different ways of enforcing satisfaction of rules.
   One way is to let a user or a computer ``invent'' $e'$.
   A developer can specify this by assigning a rule $u$ to a role $o$.
   Any actor with role $o$
   can restore invariance of rule $u$ by inserting or deleting the right triples.
   For this reason, every rule is assigned to at least one role by means of the relation $\maintain$.
\begin{equation}
   \forall u\in\rules\ \exists o\in\roles:\ o\maintain u\label{eqn:maintain}
\end{equation}
   $o\maintain u$ means that role $o$ must keep rule $u$ satisfied, i.e.\ to keep $\sat{u}{\dataset}$ true.
   As long as rule $u$ is broken, the system should notify actors in the role(s) that are administered in $\maintain$ to do something about it.

   If $\maintain$ specifies no role for a given rule $u$, the system can prevent $e$ from having any effect.
   This implements blocking behaviour, and must be accompanied by an error message.

   Automatic enforcement is specified by the developer with a special syntax,
   the {\em enforce rule}.
   An enforce rule specifies not only the rule, but also the way to restore invariance.
   The system features an engine that restores invariance of all enforce rules without delay.
   Whenever the dataset violates an enforce rule,
   this engine will make other changes to the dataset to satisfy that rule.
   Hence, the rule is not satisfied during a small, finite amount of time.

   The system contains a dedicated role for the enforcement engine,
   so every rule has at least one role in the relation $\maintain$, means that all roles together are keeping all rules satisfied.

>>>>

   Relations serve to organize and store data, to allow a developer to represent facts.
   In this paper, variables $r$, $s$, and $d$ represent relations.
   All relations are taken from an infinite set $\Rels$.
   $\Rels$ is disjoint from $\Concepts$ and $\Atoms$.
   Every relation $r$ has a name, a source concept, and a target concept.
   The notation $r=\declare{n}{A}{B}$ denotes that relation $r$ has name $n$, source concept $A$, and target concept $B$.
   The part $\pair{A}{B}$ is called the {\em signature} of the relation.

   Specialization distributes in a straightforward way over signatures:
\begin{equation}
   \pair{A}{B}\isa\pair{C}{D}\ \Leftrightarrow\ A\isa C\ \wedge\ B\isa D
\label{eqn:isa-signature}
\end{equation}

@@

   A relation $r$ can serve as a container of pairs,
   as defined by the function $\id{pop}_r:\Dataset\rightarrow\powerset{\Pair{\Atoms}{\Atoms}}$.
   It defines a set of pairs, also known as the population of $r$:
\begin{equation}
   \pop{r}{\dataset}\ =\ \{ \pair{a}{b}\mid\ \triple{a}{r}{b}\in\triples_{\dataset}\}
\label{eqn:pop-rel}
\end{equation}
   Equation~\ref{eqn:wellTypedEdge} implies that for every dataset $\dataset$:
\[\pair{a}{b}\in\pop{\declare{n}{A}{B}}{\dataset}\ \Rightarrow\ a\inst_{\dataset}A\ \wedge\ b\inst_{\dataset}B\]
   For a developer, this means that the type of an atom depends only on the relation in which it resides; not on the actual population of the database.

   Concepts too can be seen as containers of atoms,
   defined by the function $\id{pop}_C:\Dataset\rightarrow\powerset{\Atoms}$.
\begin{equation}
   \pop{C}{\dataset}\ =\ \{ x\mid\ x\ \inst_{\dataset}\ C\}
\label{eqn:pop-concept}
\end{equation}

@@

   A schema must ensure that triples ``make sense'' in the semantics defined by the schema.
   This means that it must satisfy the following requirements.
\begin{eqnarray}
   \declare{n}{A}{B}\in\rels&\Rightarrow&A\in\concepts\ \wedge\ B\in\concepts
   \label{eqn:relationsIntroduceConcepts}\\
   A\ \isa\ B&\Rightarrow&A\in\concepts\ \wedge\ B\in\concepts
   \label{eqn:isasIntroduceConcepts}\\
   \pair{A}{B}\isa\pair{C}{D}\ \wedge\ \declare{n}{C}{D}\in\rels&\Rightarrow&
   \begin{array}[t]{@{}l}
      \declare{n}{A}{B}\in\rels\ \wedge\\
      \pop{\declare{n}{A}{B}}{\dataset}\subseteq\pop{\declare{n}{C}{D}}{\dataset}
   \end{array}
   \label{eqn:subrelations} \\
   \isa\ \text{is a partial order}
   \label{eqn:isasPartialOrder}\\
   \exists\dataset\in\Dataset-\emptyset:\ \sat{u}{\dataset}
   \label{eqn:trueifyableRules}
\end{eqnarray}
   The type system of Ampersand~\cite{vdWoude2011} ensures that every information system it generates satisfies these requirements.
   Note that these requirements do not depend on any particular dataset,
   so they can be checked at compile time.
   This is also known as static typing,
   which has well established advantages for the software engineering process~\cite{HanenbergKRTS14,Petersen2014}.

@@

   The information system enforces the semantics by ensuring that the following requirements remain true at all times:
\begin{eqnarray}
   \triple{a}{\declare{n}{A}{B}}{b}\in\triples&\Rightarrow&\declare{n}{A}{B}\in\rels
   \label{eqn:define R}\\
   \forall u\in\rules&:&\sat{u}{\dataset}
   \label{eqn:satisfaction}
\end{eqnarray}
   Requirement~\ref{eqn:define R} ensures that the information system only works with triples whose relation is defined in the schema.
   Requirement~\ref{eqn:satisfaction} ensures that the dataset satisfies all rules in the schema.
   Note that these requirements are enforced at runtime.

\subsection{Changing data}
\label{sct:Changing data}
   Every rule in the schema serves as invariant,
   i.e.\ a constraint that the data in the database must satisfy at all times.
   If an actor tries to change the dataset in a way that violates a rule,
   the system will either reject that change or help the actor to find additional changes to keep that rule satisfied.

@@

   Events can cause a dataset to change.
   So, we model events as functions $\Dataset\rightarrow\Dataset$.
   We will use variables $e$, $e'$, $e''$, and so on to represent events.
   We distinguish just two types of event: insert and delete events.
   An event is a label (\id{ins} or \id{del}) with a dataset that is to be inserted or deleted.
   If $\triples$ is the set of triples inserted or deleted by an event $e$,
   we denote that event either by $\ins{\triples}$ or $\del{\triples}$.
   Equation~\ref{eqn:event} defines the effect of an event on a dataset.
\begin{equation}
   \begin{array}{rcl}
      e(\dataset)&=&\text{case}\ e\ \text{of}\\
      &&\begin{array}[t]{@{}rcl}
         \ins{\triples}&=&\pair{\triples_{\dataset}\cup\triples}{\id{insInst}_{\triples}(\inst_{\dataset})}\\
         \del{\triples}&=&\pair{\triples_{\dataset}-\triples}{\id{delInst}_{\triples}(\inst_{\dataset})}
      \end{array}
   \end{array}
   \label{eqn:event}
\end{equation}
   in which $\id{insInst}_{\triples}$ transforms $\inst_{\dataset}$ such that $\ins{\triples}(\dataset)$ complies with
   equations~\ref{eqn:specialization} and~\ref{eqn:wellTypedEdge},
   and likewise for $\id{delInst}_{\triples}$.

   As the system registers events, it inserts or deletes triples in its dataset.
   If such an event would violate requirement~\ref{eqn:satisfaction},
   either the system stays in the state $\dataset$ or there is a reaction $e'$ that restores invariance of the affected rules:
\begin{equation}
   \sat{u}{\dataset}\Rightarrow\sat{u}{e'(e(\dataset))}
   \label{eqn:restoring}
\end{equation}
   In this way, the system can restore invariance of rules and keep requirement~\ref{eqn:satisfaction} satisfied.
   The combination of events $e$ and $e'$ is called a {\em transaction}.

   In practice, there are different ways of enforcing satisfaction of rules: automatic enforcement and manual enforcement.

   Automatic enforcement is specified by the developer with a special syntax:
   the {\em enforce rule}.
   An enforce rule specifies not only the rule,
   but it implicitly defines event $e'$ to restore the invariance.
   The system must have an engine that restores invariance of all enforce rules,
   so users will not notice any violations.

   Manual enforcement means that the system (temporarily) allows violations,
   to let the user ``invent'' a reaction $e'$ that restores invariance.
   For this purpose, Ampersand relaxes requirement~\ref{eqn:satisfaction} as follows:
\begin{eqnarray}
   \forall u\in\rules&:&(\exists o\in\roles:\ o\maintain u)\ \vee\ \sat{u}{\dataset}
   \label{eqn:satisfaction relaxed}
\end{eqnarray}
   The purpose of this requirement is to allow the compiler to generate code for users with role $o$,
   to restore invariance.
   In the Ampersand language, a developer explicitly assigns a rule $u$ to a role $o$,
   to keep control over the type of persons that are allowed to ``break rules temporarily''.
   An actor, who attempts an event that violates a rule, should receive notification of the system with an appropriate error message.

   If a rule $u$ is not assigned to any role and it is not an enforce rule,
   the system must prevent any event $e$ from having any effect on the database.
   So, if $e$ would violate rule $u$, the system must reject $e$.
   This implements blocking behaviour, and must be accompanied by an error message.
